---
title: "Data Visualization Lab"
subtitle: "IPS Team"
output: rmdformats::downcute
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Complete the following lab as a group. This document should exist in your GitHub repo while you're working on it. Your code should be heavily commented so someone reading your code can follow along easily. See the first code snippet below for an example of commented code.**

**Here's the catch: For any given problem, the person writing the code should not be the person commenting that code, and every person must both code AND comment at least one problem in this lab (you decide how to split the work). This will involve lots of pushing and pulling through Git, and you may have to resolve conflicts if you're not careful! Refer to last Thursday's class notes for details on conflict resolution.**

**ALSO, all plots generated should have labeled axes, titles, and legends when appropriate. Don't forget units of measurement! Make sure these plots could be interpreted by your client.**


# Playing with the indoor positioning system data

The `IPS_sampledata` data set contains a fraction of the indoor positioning system data for 15 randomly sampled locations.This data set is loaded into memory using the chunk of R code below, complete the following exercises. 

The exercises below aim to help you polish your data visualization skills in R while simulatanelously helping you gain a better understanding of the IPS data.

First, let's load up the data set.

```{r eval=TRUE, echo=TRUE, message=FALSE}
# loads data set IPS_sampledata
load("IPS_portion.RData")
write.csv(IPS_sampledata,"IPS_sampledata")
```

```{r}
#install packages
install.packages("dplyr")
library(dplyr)
library(magrittr)
```

## Variable dictionary

- `time`: timestamp in milliseconds since midnight 01/01/1970 UTC

- `scanMac`: MAC address of the scanning device (this is a handheld device)

- `posX`, `posY` and `posZ`: the (x, y, z) physical coordinate of the scanning device

- `orientation`: degree orientation of the user carrying the scanning device in degrees

- `mac`: MAC address of an access point

- `signal`: signal strength in dBm (Decibel-milliwatts)

- `channel`: the channel frequency

- `type`: type of device (access point = 3, device in adhoc mode = 1)

## Preparing the data

For this part of the lab you may use some of the code from the previous two labs to save some time.

1. Transform the variables in the data set into their correct types.

```{r transform_vars}
# transform variables into correct types
IPS_sampledata <- IPS_sampledata %>%
  mutate(across(c(time, posX, posY, posZ, orientation, signal, channel, type), ~ suppressWarnings(as.numeric(.))))
# Creates a function to convert the observations in the time variable from milliseconds to into the POSIXct time format 
posixct_func<- function(x){  
  seconds<- x %/% 1000
  date_time<-as.POSIXct(seconds, origin = "1970-01-01", tz = "GMT")
}
# Creates a new data frame with with time converted into posixct format
IPS_sampledata_change_time <-IPS_sampledata %>% mutate(time = posixct_func(time))

IPS_sampledata_change_time
```


2. From `IPS_sampledata`, keep only records for the six access points included in the `mac_locs` table, save these to `IPS_sampledata`.

```{r filter_by_macs}
#setwd("C:/Users/risin/OneDrive/Desktop/School/Classes/STAT410/Git Repos/Codinglab/Codinglab/Codinglab/Codinglab/Indoor-Positioning-System/Data-Visualization/Data-Visualization/Data-Visualization/Lab03")
mac_locs <- readr::read_table("accessPointLocations.txt")

MAC_addresses<-c(mac_locs$Macs)

IPS_sampledata_mac <- IPS_sampledata_change_time %>% filter(mac %in% MAC_addresses)
IPS_sampledata_mac
```

3. Create a grouped version of the `orientation` variable, with groups (bins) determined by the proximity to angles 0, 45, 90, 135, 180, 225, 270, 315.

```{r bin_orientation}
# This function takes and an angle and locates its closest proximity to angles in the bin vector
nearest_angle <- function(angle, bins) {
  bins[which.min(abs(angle %% 360 - bins))]
}
bins <- c(0, 45, 90, 135, 180, 225, 270, 315)

# Applies the nearest_angle function to the orientation column and creates a group version of the orientation variable conformed to angles 0, 45, 90, 135, 180, 225, 270, 315.
IPS_sampledata_orientation <- IPS_sampledata_mac %>%
mutate(rec_orient = sapply(orientation, nearest_angle, bins = bins))
head(IPS_sampledata_orientation)
tail(IPS_sampledata_orientation)
```


4. Combine the information in `IPS_sampledata` and `mac_locs` using one of the `join_` functions we learned in class.

```{r mac_joins}
# rename the Macs column to match the data set
colnames(mac_locs)[colnames(mac_locs) == "Macs"] <- "mac"
# use inner join to find the matched data
IPS_sampledata_joined <- IPS_sampledata_orientation %>% inner_join(mac_locs, by = c("mac"))
head(IPS_sampledata_joined)
```


## Getting useful visualizations

Throughout this section you'll have some freedom to create useful visualizations for your project.  [**Here is a link**](https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf) to a `ggplot2` cheat-sheet, and [**here is another link**](https://r-graph-gallery.com/index.html) to a nice R graph gallery for inspiration.

1. Generate a scatter plot showing the locations (`posX` and `posY` in `IPS_Sampledata`) of the handheld device, as well as those from the access points (variables `x` and `y` from `mac_locs`).   Use the `aesthetic` of your choosing to distinguish the handheld device locations from those corresponding to the access points.

```{r loc_scatter}
library(ggplot2)
unique_combinations <- IPS_sampledata  %>% distinct(posX, posY)

combined_cordinates <- bind_rows(
  unique_combinations %>% rename(X = posX, Y = posY),  
  mac_locs %>% rename(X = x, Y = y)) %>% select(X,Y)  

combined_cordinates_with_groups <- combined_cordinates %>%
  mutate(coordinates_of = ifelse(row_number() > 15, "Access_Points", "Locations"))
head(combined_cordinates_with_groups,n=21)

ggplot(combined_cordinates_with_groups, aes(x = X, y = Y , color = coordinates_of)) +
  geom_point() +
  scale_color_manual(values = c("Locations" = "black", "Access_Points" = "red"))
```



2. Use the resulting data set from exercise 4 from the *"Preparing the Data"* section, `group` and `summarize` it in the appropriate way to obtain the median, mean, standard deviation and any other summary statistic you consider useful to assess the signal strength for each combination of:  location, orientation (the binned version), and access point.

```{r summary_by_combo}

IPS_summary <- IPS_sampledata_joined %>%
  group_by(posX, posY, rec_orient, mac,x,y) %>%
  summarise(mean_signal = mean(signal, na.rm = TRUE),
    median_signal = median(signal, na.rm = TRUE),
    sd_signal = sd(signal, na.rm = TRUE),
    min_signal = min(signal, na.rm = TRUE),
    max_signal = max(signal, na.rm = TRUE),
    IQR_signal = IQR(signal, na.rm = TRUE),
    count = n(),
    .groups = "drop"  # Removes grouping from the resulting data frame
  )
print(IPS_summary,n = 50, width = Inf)
```


3. Using the data generated from the exercise above, produce one or more visualizations (at most 3) to explore the relationship between  orientation angle, location (relative to the access point), and signal strength, using `ggplot2` tools discussed in class. Briefly comment on the relationships you observe in your visualization(s) between orientation angle, distance between locations and access points, and signal strength. From your visualization(s), what cardinal direction can you relate the 0 degree orientation with?

```{r cool_dat_vis}
# your code here
IPS_dist<-IPS_summary
dist <- sqrt((IPS_dist$posX - IPS_dist$x)^2 + (IPS_dist$posY - IPS_dist$y)^2)
dist <- round(dist, digits = 2)
IPS_dist <- data.frame(append(IPS_dist, list(dist = dist), after = 6))
head(IPS_dist)
```

```{r cool_dat_vis}
IPS_angle<-IPS_dist

# Calculate the angle in radians from device to router
angle <- atan2(IPS_angle$y - IPS_angle$posY, IPS_angle$x - IPS_angle$posX)

# If desired, convert the angle to degrees
angle <- angle * (180 / pi)

# Adjust to ensure all angles are in the range 0-359 degrees
angle <- (angle + 360) %% 360

# Round to the 6th decimal place
angle <- round(angle, digits = 2)

IPS_angle <- data.frame(append(IPS_angle, list(angle = angle), after = 3))
head(IPS_angle,48)

```

```{r}
result1 <- IPS_angle %>%
  group_by(posX, posY, mac, rec_orient) %>%
  # Step 2: Calculate the angular difference, considering the circular nature
  mutate(
    angle_diff = abs(rec_orient - angle),
    # Correct for values greater than 180 degrees by subtracting from 360
    angle_diff = ifelse(angle_diff > 180, 360 - angle_diff, angle_diff)) %>% 
  select(
    1:4,           # select the first 4 columns to be the first
    angle_diff,    # moves angle_diff to the 5th column spot      
    everything()   # Then adds the rest of the columns in order
  )
print(result1, n = 30)

# Checks to make sure there are no angle differences below zero or above 180, or NA values 
unique_angle_diff<-unique(result1$angle_diff)
# Logical condition for values below 0, above 180, or NA
invalid_cases <- unique_angle_diff < 0 | unique_angle_diff > 180 | is.na(unique_angle_diff)
# Extract the problematic values
problem_values <- unique_angle_diff[invalid_cases]
print(problem_values)
```


```{r}
typeof(combined_cordinates_with_groups) # Identifies the type/structure we will be working with

data4_use1<-combined_cordinates_with_groups # Copy the structure and data

data4_use1<- data4_use1[c(2,17,19,20), ]  # Subset rows 2,17,19, and 20

ggplot(data4_use1, aes(x = X, y = Y , color = coordinates_of)) +
  geom_point() +
  
  scale_color_manual(values = c("Locations" = "black", "Access_Points" = "red")) + 
  scale_x_continuous(limits = c(0, 35)) +
  scale_y_continuous(limits = c(-3, 15))
```


```{r}
filtered_data <- result1 %>% 
  ungroup() %>%  # Remove grouping
  slice(c(5,18,40)) %>% select(posX,posY,rec_orient,angle,angle_diff,mac,x,y,dist,mean_signal)
print(filtered_data)
```


```{r}
result2 <- result1 %>% group_by(posX,posY,mac) %>% summarise(
    angle_diff = min(angle_diff),  # Find the minimum value for the angle difference
    .groups = "drop"  # grouping is not necessary at this point
  )
print(result2)
```

```{r}
result3<- result1 %>% semi_join(result2)
print(result3)
```

```{r}
ggplot(result3, aes(x = dist, y = mean_signal)) +
  geom_point() +
  facet_wrap(.~mac) +
  geom_smooth()
```

```{r}
ggplot(result1, aes(x = dist, y = mean_signal, color = angle_diff)) +
  geom_point() +
  facet_wrap(.~mac) +
  geom_smooth()
```

```{r}

IPS_position_coordinates2 <- result1 %>%
  mutate(pos_coord = paste0("(",posX,",",posY,")")) %>% 
  select(
    1:2,           # select the first 2 columns to be the first
    pos_coord,    # moves pos_coord to the 3rd column spot     
    everything()   # Then adds the rest of the columns in order
  )

ggplot(IPS_position_coordinates2, aes(x = angle_diff, y = mean_signal)) +
  geom_point() +
  facet_grid(mac~pos_coord)
```

4. Replicate the plot attached in the link.

```{r rep_figure}
#xfun::embed_file("SignalByLocation.pdf")
IPS_position_coordinates <- IPS_sampledata_joined %>%
  mutate(pos_coord = paste0("(",posX,",",posY,")")) %>% 
  select(
    1:2,           # select the first 2 columns to be the first
    pos_coord,    # moves pos_coord to the 3rd column spot     
    everything()   # Then adds the rest of the columns in order
  )

print(IPS_position_coordinates, n = 36, quote = FALSE)

ggplot(IPS_position_coordinates, aes(x = pos_coord, y = signal, color = pos_coord)) +
  geom_point() +
  geom_boxplot() +
  facet_grid(mac~rec_orient) +
  labs(
    title = "Signal Strength by Location\n (by access point/orientation combination)",
    x = "location coordinates",
    y = "signal",
  ) +
  theme(axis.text.x = element_text(angle = 90))# Set angle to 90 degrees (Vertical)
```
